Parsing Postfix log files

### 2007-11-11 doc/logparser.tex
More corrections. Capitalised Postfix in the correct places. Adopted a standard formatting for postfix/foo daemons.

@@

    \item [program] The program (\texttt{postfix/smtpd},
        \texttt{postfix/qmgr}, etc.) whose log lines the rule applies to.
        This avoids needlessly trying rules which won't match the line, or
        worse, might match unintentionally.  Rules whose program is
        \texttt{*} will be tried against any lines which aren't parsed by
        program specific rules.

@@

\textbf{Field}      & \textbf{Value}                                    \\
name                & Unknown sender domain                             \\
description         & We do not accept mail from unknown domains        \\
restriction\_name   & reject\_unknown\_sender\_domain                   \\
postfix\_action     & REJECTED                                          \\
program             & postfix/smtpd                                     \\
regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
                    & \verb!Sender address rejected: Domain not found;! \\
                    & \verb!from=<\5> to=<(__RECIPIENT__)> !            \\
                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
result\_cols        & recipient = 6; sender = 5                         \\
connection\_cols    & helo = 7                                          \\
result\_data        &                                                   \\
connection\_data    &                                                   \\
action              & REJECTION                                         \\
queueid             & 1                                                 \\
hits                & 0                                                 \\
hits\_total         & 0                                                 \\
priority            & 0                                                 \\

@@

    \item Mail enters the system via \SMTP{} or local submission.

@@

    \item Follow the progress of the accepted mail until it's either
        delivered, bounced or deleted, log all data, and finish.

@@

The most difficult complication initially encountered is that locally
addressed mails are not always delivered directly to a mailbox: sometimes
they are accepted for a local address but need to be delivered to one or
more remote addresses due to aliases.  When this occurs a child mail will
be injected into the postfix queue, but without the explicit logging smtpd
or sendmail injected mails have.  Thus the source is not immediately
discernible from the log line in which the mail first appears; from a
strictly chronological reading of the logs it \textit{usually\/} appears as
if the child mail has appeared from thin air.  Subsequently the parent mail
will log the creation of the child mail:

@@

Unfortunately while all log lines from an individual process appear in
chronological order, the order in which log lines from different processes
are interleaved is subject to the vagaries of process scheduling.  In
addition the first log line belonging to the child mail (the log line cited
above properly belongs to the parent mail) is logged by qmgr,\footnote{Qmgr
is the Postfix daemon which manages the mail queue, determining which mails
will have delivery attempted next.} so the order also depends on how busy
qmgr is.\footnote{Postfix is quite paranoid about mail delivery, an
excellent characteristic for an MTA to possess, so it won't log that the
child has been created until it is absolutely certain that the mail has
been written to disk.}

Because of this the parser cannot complain when it encounters a log line
from qmgr for a previously unseen mail; it must flag the mail as coming
from an unknown origin, and subsequently clear the flag if and when the
origin of the mail becomes clear.  Obviously the parser could omit checking
of where mails originate from, but the author believes that it is better to
require an explicit source, as bugs are more likely to be exposed.

Process scheduling can have a still more confusing effect: quite often the
child mail will be created, delivered and entirely finished with
\textbf{before} the parent logs the creation line!  Thus mails flagged as
coming from an unknown origin cannot be entered into the database when
their final log line is parsed; instead they must be marked as ready for
entry and subsequently entered by the parent mail once it has been
identified.  Quite apart from identifying mail injected in an unknown
fashion, or bugs in the parser, unknown origin mails need to be marked as
such because they lack some data present in the parent mail; they must copy
that data from their parent before being entered in the database.  The
effect of this complication upon the algorithm is discussed in
section~\refwithpage{tracking re-injected mail}.

@@

Section~\refwithpage{actions-in-detail} explains the actions in substantive
detail; this section will omit such detail because it would clutter and
confuse the algorithm description.  The data flow chart in
section~\refwithpage{flow-chart} should also be consulted while reading
this section.  For the sake of clarity this description covers the most
common paths through the algorithm, as including every path would hinder
understanding.  The more uncommon paths are caused by the complications
described in section~\refwithpage{additional complications}, and are
covered therein.

@@

Everything starts off with a mail entering the system, whether by local
submission via postdrop/sendmail, by \SMTP{}, by re-injection due to
forwarding, or internally generated by Postfix.  Local submission is the
simplest case: a queueid is assigned immediately and the sender address is
logged (action: pickup; flowchart:~2).

@@

The obvious counterpart to mail entering the system is mail leaving the
system, whether it is by deletion, bouncing, or local or remote delivery.
All four are handled in exactly the same way:

@@

    \item The sender and recipient addresses will be logged separately
        (action: save\_by\_queueid; flowchart:~10).

@@

It should be reiterated that the actions above happen whether the mail is
delivered to a mailbox, piped to a command, delivered to a remote server,
bounced (due to a mail loop, delivery failure, or five day timeout), or
deleted by the administrator.  The exception is what happens after delivery
to the parent or children of mail re-injected due to forwarding, and that
is explained in section~\refwithpage{tracking re-injected mail}.

@@

    \item The indicator which more commonly introduces re-injection is when
        \texttt{postfix/qmgr} selects a mail with a previously unseen
        queueid for delivery (action: mail\_picked\_for\_delivery;
        flowchart:~3), in which case a new data structure will be created.
        The mail will be flagged as having unknown origins; this flag
        should be subsequently cleared once the origin has been
        established.  This may also be an indicator that the mail is a
        bounce notification, see
        section~\refwithpage{identifying-bounce-notifications} for details.

    \item Local delivery re-injects the mail and logs a relayed delivery
        rather than delivering directly to a mailbox or program as it
        usually would (action: track; flowchart:~11).\footnote{Relayed
        delivery is performed by the \SMTP{} client; local delivery means
        local to the server, i.e.\ an address the server is final
        destination for.} In this case the mail may already have been
        created (action: qmgr\_chooses\_mail; flowchart:~3; described
        above) and the unknown origin flag will be cleared; if not a new
        data structure will be created.  In both cases the new mail is
        marked as a child of the parent.  The log line in question is:

@@

        This second indicator always occurs for re-injected mail but
        typically after the first indicator explained above.  This
        indicator is required to tie the parent and child mails together
        and so is central to the process of tracking re-injected mails.

@@

    \item [REJECTION] Deal with postfix rejecting an \SMTP{} command from
        the remote client: log the rejection with a mail if there is a
        queueid in the log line, or with the connection if not.

@@

Occasionally a log file will have out of order log lines which cannot be
dealt with by the techniques described in sections~\refwithpage{tracking
re-injected mail},~\refwithpage{discarding cleanup
lines}~and~\refwithpage{pickup logging after cleanup}.  In the 93 log files
used for testing this occurs only five times in 60,721,709 log lines, but
for completeness of the algorithm it should be dealt with.  The five
occurrences in the test log files have the same characteristics: the
\texttt{postfix/local} log line showing delivery to a local mailbox occurs
after the \texttt{postfix/qmgr} log line showing removal of the mail from
the queue because delivery is completed.  This causes problems: the mail is
not complete, so entry into the database fails; a new mail is created when
the \texttt{postfix/local} line is parsed and remains in the state tables;
four warnings are issued per pair of out of order log lines.

@@

The list of valid combinations is given below.  Every mail will
additionally have log entries from \texttt{postfix/cleanup} and
\texttt{postfix/qmgr}; any mail may also have log entries from
\texttt{postfix/bounce}, \texttt{postfix/postsuper}, or both.

@@

    \item [postfix/local:] Local delivery of bounce notification, or local
        delivery of forwarded/tracked mail.

    \item [postfix/local, postfix/pickup:] Locally submitted mail, locally
        delivered.

    \item [postfix/local, postfix/pickup, postfix/smtp:] Locally submitted
        mail, \newline both local and remote delivery.

    \item [postfix/local, postfix/smtp, postfix/smtpd:] Mail accepted from
        a remote client, both local and remote delivery.

    \item [postfix/local, postfix/smtpd:] Mail accepted from a remote
        client, local delivery.

    \item [postfix/pickup, postfix/smtp:] Locally submitted mail, remote
        delivery.

    \item [postfix/smtp:] Remote delivery of forwarded/tracked mail.

    \item [postfix/smtp, postfix/smtpd:] Mail accepted from a remote
        client, remotely delivered (relaying mail for clients on the local
        network).

@@

The aborted delivery attempts described in
section~\refwithpage{aborted-delivery-attempts} occur frequently, but the
aborted delivery attempts described in this section only occur four times
in the 93 log files used for testing.  The symptoms are the same as in
section~\refwithpage{aborted-delivery-attempts}, except that there
\textit{is\/} a \texttt{postfix/cleanup} log line; there does not appear to
be anything in the log file to explain why there are no further log
messages.  The only way to detect these mails is to periodically scan all
mails in the state tables, deleting any mails displaying the following
characteristics:

@@

    \item There must be exactly two \texttt{postfix/smtpd} and one
        \texttt{postfix/cleanup} log entries for the mail, with no
        additional log entries.

@@

    \item Postfix delivers the mail to the local mailbox and
        username@domain, as instructed by the user's forwarding
        instructions. The forwarded mail has a
        \texttt{Delivered-To:~username@domain} header added, and the
        envelope sender address is username@domain.  Log lines are added by
        \textit{postfix/local}, \textit{postfix/qmgr\/} (twice),
        \textit{postfix/cleanup\/} and finally \textit{postfix/pickup}.

    \item Postfix accepts the mail for username@domain, but while
        delivering it notices that the \texttt{Delivered-To} header already
        contains the address it's currently delivering to, and therefore
        sends a bounce notification with sender address \textit{$<>$\/} to
        the original sender: username@domain.  Log lines are added by
        \textit{postfix/local}, \textit{postfix/qmgr\/} (twice) and
        \textit{postfix/cleanup}.

    \item Postfix delivers the bounce notification to both the local
        mailbox and to username@domain, as instructed by the user's
        forwarding instructions.  A \texttt{Delivered-To:~username@domain}
        header is added to the forwarded bounce notification, and the
        envelope sender address is now username@domain.  Log lines are
        added by \textit{postfix/local}, \textit{postfix/qmgr\/} (twice),
        \textit{postfix/cleanup}, and finally \textit{postfix/pickup}.

    \item Postfix accepts the forwarded bounce notification but while
        delivering it notices that the \texttt{Delivered-To} header already
        contains the address currently being delivered to, and sends a
        bounce notification to the sender: username@domain.  Log lines are
        added by \textit{postfix/local}, \textit{postfix/qmgr\/} (twice)
        and \textit{postfix/cleanup}.

@@

The vast majority of log lines when a mail loop occurs are from Postfix
components which have a small number of rules associated with them, whereas
in general the most common log lines are logged by \textit{postfix/smtpd},
which also has the highest number of rules.  \textit{postfix/smtpd\/} log
lines are distributed across rules much more evenly that
\textit{postfix/qmgr}, \textit{postfix/local}, \textit{postfix/cleanup\/}
or \textit{postfix/pickup}, so the average number of rules required to
parse a \textit{postfix/smtpd\/} log line is much higher that the average
number required to parse other log lines.  Thus the average number of rules
required to parse a random line from logs containing a mail loop is much
lower than the average number required to parse a random log line from logs
without a mail loop, as can be seen in the peaks in
graphs~\refwithpage{execution time vs file size vs number lines
factor},~\refwithpage{percentage increase of shuffled over normal}
and~\refwithpage{percentage increase of reversed over normal}.  The
difference is more extreme in the reversed ordering case because the
largest number of rules to be consulted for the commonly occurring log
lines during a mail loop is far less than the largest number of rules that
are consulted for commonly occurring lines outside of a mail loop.  The
number of rules which needs to be consulted when the ordering is shuffled
varies between the optimum and nadir, and the performance varies
proportionally.

>>>>

% A command to format a Postfix daemon's name
\newcommand{\daemon}[1]{%
    \texttt{postfix/#1}%
}

@@

    \item [program] The program (\daemon{smtpd}, \daemon{qmgr}, etc.) whose
        log lines the rule applies to.  This avoids needlessly trying rules
        which won't match the line, or worse, might match unintentionally.
        Rules whose program is \texttt{*} will be tried against any lines
        which aren't parsed by program specific rules.

@@

\textbf{Field}      & \textbf{Value}                                    \\
name                & Unknown sender domain                             \\
description         & We do not accept mail from unknown domains        \\
restriction\_name   & reject\_unknown\_sender\_domain                   \\
postfix\_action     & REJECTED                                          \\
program             & \daemon{smtpd}                                    \\
regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
                    & \verb!Sender address rejected: Domain not found;! \\
                    & \verb!from=<\5> to=<(__RECIPIENT__)> !            \\
                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
result\_cols        & recipient = 6; sender = 5                         \\
connection\_cols    & helo = 7                                          \\
result\_data        &                                                   \\
connection\_data    &                                                   \\
action              & REJECTION                                         \\
queueid             & 1                                                 \\
hits                & 0                                                 \\
hits\_total         & 0                                                 \\
priority            & 0                                                 \\

@@

    \item Mail enters the system via \SMTP{} or local submission; and new
        data structure is created for it.

@@

    \item Follow the progress of the accepted mail until it's either
        delivered, bounced or deleted, then log all data, and finish.

@@

The most difficult complication initially encountered is that locally
addressed mails are not always delivered directly to a mailbox: sometimes
they are addressed to and accepted for a local address but need to be
delivered to one or more remote addresses due to aliases.  When this occurs
a child mail will be injected into the Postfix queue, but without the
explicit logging smtpd or postdrop injected mails have.  Thus the source is
not immediately discernible from the log line in which the mail first
appears; from a strictly chronological reading of the logs it
\textit{usually\/} appears as if the child mail has appeared from thin air.
Subsequently the parent mail will log the creation of the child mail:

@@

Unfortunately while all log lines from an individual process appear in
chronological order, the order in which log lines from different processes
are interleaved is subject to the vagaries of process scheduling.  In
addition the first log line belonging to the child mail (the log line cited
above properly belongs to the parent mail) is logged by qmgr,\footnote{Qmgr
is the Postfix daemon which manages the mail queue, determining which mails
will have delivery attempted next.} so the order also depends on how busy
qmgr is.\footnote{Postfix is quite paranoid about mail delivery, an
excellent characteristic for an \MTA{} to possess, so it won't log that the
child has been created until it is absolutely certain that the mail has
been written to disk.}

Because of this the parser cannot complain when it encounters a log line
from qmgr for a previously unseen mail; it must flag the mail as coming
from an unknown origin, and subsequently clear the flag if and when the
origin of the mail becomes clear.  Obviously the parser could omit checking
of where mails originate from, but the author believes that it is better to
require an explicit source, as bugs in the parser are more likely to be
exposed.

Process scheduling can have a still more confusing effect: quite often the
child mail will be created, delivered and entirely finished with
\textbf{before} the parent logs the creation line!  Thus mails flagged as
coming from an unknown origin cannot be entered into the database when
their final log line is parsed; instead they must be marked as ready for
entry and subsequently entered by the parent mail once it has been
identified.  Quite apart from identifying mail injected in an unknown
fashion, or bugs in the parser, unknown origin mails need to be marked as
such because they lack some data present in the parent mail; they must copy
that data from their parent before being entered in the database.  The
effect of this complication upon the algorithm is discussed more fully in
section~\refwithpage{tracking re-injected mail}.

XXX FIGURE OUT IF I NEED TO COPY DATA ANY MORE, AND DOCUMENT IT IF SO\@.

@@

Section~\refwithpage{actions-in-detail} explains the actions in substantive
detail; this section omits such detail because it would clutter and confuse
the algorithm description.  The flow chart in
section~\refwithpage{flow-chart} should also be consulted while reading
this section.  For the sake of clarity this description covers only the
most common paths through the algorithm, as including every path would
hinder understanding; the more uncommon paths are caused by the
complications described in section~\refwithpage{additional complications},
and are covered therein.

@@

Everything starts off with a mail entering the system, whether by local
submission via postdrop/sendmail,\footnote{Postfix provides a command that
is compatible with the Sendmail (\url{http://www.sendmail.org/ XXX CHECK})
mail submission program which all Unix commands which send mail depend on.}
by \SMTP{}, by re-injection due to forwarding, or internally generated by
Postfix.  Local submission is the simplest case: a queueid is assigned
immediately and the sender address is logged (action: pickup;
flowchart:~2).

@@

The obvious counterpart to mail entering the system is mail leaving the
system, whether by deletion, bouncing, local delivery, or remote delivery.
All four are handled in exactly the same way:

@@

    \item Postfix will log the sender and recipient addresses separately
        (action: save\_by\_queueid; flowchart:~10).

@@

It should be reiterated that the actions above happen whether the mail is
delivered to a mailbox, piped to a command, delivered to a remote server,
bounced (due to a mail loop, delivery failure, or five day timeout), or
deleted by the administrator.  The exception is what happens after delivery
to the parent or children of mail re-injected due to forwarding, as
explained in section~\refwithpage{tracking re-injected mail}.

@@

    \item The indicator which more commonly introduces re-injection is when
        \daemon{qmgr} selects a mail with a previously unseen
        queueid for delivery (action: mail\_picked\_for\_delivery;
        flowchart:~3), in which case a new data structure will be created.
        The mail will be flagged as having unknown origins; this flag
        should be subsequently cleared once the origin has been
        established.  This may also be an indicator that the mail is a
        bounce notification, see
        section~\refwithpage{identifying-bounce-notifications} for details.

    \item Local delivery re-injects the mail and logs a relayed delivery
        rather than delivering directly to a mailbox or program as it
        usually would (action: track; flowchart:~11).\footnote{Relayed
        delivery is performed by the \SMTP{} client; local delivery means
        local to the server, i.e.\ an address the server is final
        destination for.} In this case the mail may already have been
        created (described above) and the unknown origin flag will be
        cleared; if not a new data structure will be created.  In both
        cases the new mail is marked as a child of the parent.  The log
        line in question is:

@@

        This second indicator always occurs for re-injected mail but
        typically occurs after the first indicator explained above.  This
        indicator is required to tie the parent and child mails together
        and so is central to the process of tracking re-injected mails.

@@

    \item [REJECTION] Deal with Postfix rejecting an \SMTP{} command from
        the remote client: log the rejection with a mail if there is a
        queueid in the log line, or with the connection if not.

@@

Occasionally a log file will have out of order log lines which cannot be
dealt with by the techniques described in sections~\refwithpage{tracking
re-injected mail},~\refwithpage{discarding cleanup
lines}~and~\refwithpage{pickup logging after cleanup}.  In the 93 log files
used for testing this occurs only five times in 60,721,709 log lines, but
for completeness of the algorithm it should be dealt with.  The five
occurrences in the test log files have the same characteristics: the
\daemon{local} log line showing delivery to a local mailbox occurs after
the \daemon{qmgr} log line showing removal of the mail from the queue
because delivery is completed.  This causes problems: the mail is not
complete, so entry into the database fails; a new mail is created when the
\daemon{local} line is parsed and remains in the state tables; four
warnings are issued per pair of out of order log lines.

@@

The list of valid combinations is given below.  Every mail will
additionally have log entries from \daemon{cleanup} and \daemon{qmgr}; any
mail may also have log entries from \daemon{bounce}, \daemon{postsuper}, or
both.

@@

    \item [\daemon{local}:] Local delivery of bounce notification, or local
        delivery of forwarded/tracked mail.

    \item [\daemon{local}, \daemon{pickup}:] Mail submitted locally on the
        server, delivered locally on the server.

    \item [\daemon{local}, \daemon{pickup}, \daemon{smtp}:] Locally
        submitted mail, \newline both local and remote delivery.

    \item [\daemon{local}, \daemon{smtp}, \daemon{smtpd}:] Mail accepted
        from a remote client, both local and remote delivery.

    \item [\daemon{local}, \daemon{smtpd}:] Mail accepted from a remote
        client, local delivery.

    \item [\daemon{pickup}, \daemon{smtp}:] Locally submitted mail, remote
        delivery.

    \item [\daemon{smtp}:] Remote delivery of forwarded/tracked mail.

    \item [\daemon{smtp}, \daemon{smtpd}:] Mail accepted from a remote
        client, then remotely delivered (typically relaying mail for
        clients on the local network).

@@

The aborted delivery attempts described in
section~\refwithpage{aborted-delivery-attempts} occur frequently, but the
aborted delivery attempts described in this section only occur four times
in the 93 log files used for testing.  The symptoms are the same as in
section~\refwithpage{aborted-delivery-attempts}, except that there
\textit{is\/} a \daemon{cleanup} log line; there does not appear to be
anything in the log file to explain why there are no further log messages.
The only way to detect these mails is to periodically scan all mails in the
state tables, deleting any mails displaying the following characteristics:

@@

    \item There must be exactly two \daemon{smtpd} and one \daemon{cleanup}
        log entries for the mail, with no additional log entries.

@@

    \item Postfix delivers the mail to the local mailbox and
        username@domain, as instructed by the user's forwarding
        instructions. The forwarded mail has a
        \texttt{Delivered-To:~username@domain} header added, and the
        envelope sender address is username@domain.  Log lines are added by
        \daemon{local}, \daemon{qmgr} (twice), \daemon{cleanup} and finally
        \daemon{pickup}.

    \item Postfix accepts the mail for username@domain, but while
        delivering it notices that the \texttt{Delivered-To} header already
        contains the address it's currently delivering to, and therefore
        sends a bounce notification with sender address \textit{$<>$\/} to
        the original sender: username@domain.  Log lines are added by
        \daemon{local}, \daemon{qmgr} (twice) and \daemon{cleanup}.

    \item Postfix delivers the bounce notification to both the local
        mailbox and to username@domain, as instructed by the user's
        forwarding instructions.  A \texttt{Delivered-To:~username@domain}
        header is added to the forwarded bounce notification, which now has
        an envelope sender address of username@domain.  Log lines are added
        by \daemon{local}, \daemon{qmgr} (twice), \daemon{cleanup}, and
        finally \daemon{pickup}.

    \item Postfix accepts the forwarded bounce notification but while
        delivering the mail it notices that the \texttt{Delivered-To}
        header already contains the address currently being delivered to,
        and sends a bounce notification to the sender: username@domain.
        Log lines are added by \daemon{local}, \daemon{qmgr} (twice) and
        \daemon{cleanup}.

@@

The vast majority of log lines when a mail loop occurs are from Postfix
components which have a small number of rules associated with them, whereas
in general the most common log lines are logged by \daemon{smtpd}, which
also has the highest number of rules.  \daemon{smtpd} log lines are
distributed across rules much more evenly that \daemon{qmgr},
\daemon{local}, \daemon{cleanup} or \daemon{pickup}, so the average number
of rules required to parse a \daemon{smtpd} log line is much higher that
the average number required to parse other log lines.  Thus the average
number of rules required to parse a random line from logs containing a mail
loop is much lower than the average number required to parse a random log
line from logs without a mail loop, as can be seen in the peaks in
graphs~\refwithpage{execution time vs file size vs number lines
factor},~\refwithpage{percentage increase of shuffled over normal}
and~\refwithpage{percentage increase of reversed over normal}.  The
difference is more extreme in the reversed ordering case because the
largest number of rules to be consulted for the commonly occurring log
lines during a mail loop is far less than the largest number of rules that
are consulted for commonly occurring lines outside of a mail loop.  The
number of rules which needs to be consulted when the ordering is shuffled
varies between the optimum and nadir, and the performance varies
proportionally.
